Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-20 optim.qmd
intro

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *projection
pursuit index (PPI)*, (@hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020). Given
high-dimensional data $X \in \mathcal{R}^{n }\times p$ and PPI
$f(\cdot)$, PP finds the orthonormal projection matrix
$A \in \mathcal{R}^{p \times d}$ by solving the following optimisation
problem:

$$
\underset{A}{\max} f(XA) \quad \text{subject to} \quad A'A = I
$$

@@

The artificial jellyfish search (JS) algorithm [@chou_novel_2021] is a
swarm-based metaheuristic optimisation algorithm inspired by the search
behaviour of jellyfish in the ocean. It is one of the latest swarm
intelligence algorithms [@rajwar_exhaustive_2023], which was shown to
have stronger search ability and faster convergence with fewer
parameters compared to classic optimisation methods
[@chou_novel_2021]-[@chou_recent_2022].

The primary goal of this study is to investigate the performance of JSO in the context of the PP guided tour. A series of simulation experiments using well-recognised data sets and PPIs are conducted to yield insights regarding the behaviour of JSO under different settings and the sensitivity of hyper-parameter in the optimisation. Second, to observe the performance of JSO with different types of PPI, a collection of metrics is introduced to capture specific properties of the index including squintability and smoothness as introduced in @laa_using_2020. To the best of the author’s knowledge, this is the first attempt to quantitatively measure squintability and smoothness. Finally, the relationship between the JSO performance, hyper-parameters tuning and various properties of PPI is analysed to provide helpful guidance for practitioners that are using the guided tour for high dimensional visualisation.

The rest of this paper is structured as follows. @sec-background
introduces the PP guided tour. @sec-theory describes the details of JSO
and introduces metrics that measure different characteristics of a PPI.
@sec-simulation describes a simulation study on the performance of the
JSO using well-known data sets and index functions. @sec-conclusion
summarises the work and provides suggestions for future directions.

@@

The speed of optimizing an index function can be calculated/measured
using the computational complexity (in big O notation, with respect to
the sample size) of computing the index function.

>>>>

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely *projection
pursuit index (PPI)* (@hall1989polynomial, @cook1993projection,
@lee2010projection, @Loperfido2018, @Loperfido2020). Given
high-dimensional data $X \in \mathbf{R}^{n }\times p$ and PPI
$f(\cdot)$, PP finds the orthonormal projection matrix
$A \in \mathbf{R}^{p \times d}$ by solving the following optimisation
problem:

$$
\underset{A}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I
$$

@@

The artificial Jellyfish Search Optimiser (JSO) [@chou_novel_2021] is a
swarm-based metaheuristic optimisation algorithm inspired by the search
behaviour of jellyfish in the ocean. It is one of the latest swarm
intelligence algorithms [@rajwar_exhaustive_2023], which was shown to
have stronger search ability and faster convergence with fewer
parameters compared to classic optimisation methods
[@chou_novel_2021]-[@chou_recent_2022].

The primary goal of this study is to investigate the performance of JSO in the context of the PP guided tour. A series of simulation experiments using well-recognised data sets and PPIs are conducted to yield insights regarding the behaviour of JSO under different settings and the sensitivity of hyper-parameter in the optimisation. Second, to observe the performance of JSO with different types of PPI, a collection of metrics is introduced to capture specific properties of the index including squintability and smoothness as introduced in @laa_using_2020. To the best of the authors’ knowledge, this is the first attempt to quantitatively measure squintability and smoothness. Finally, the relationship between the JSO performance, hyper-parameters tuning and various properties of PPI is analysed to provide helpful guidance for practitioners that are using the guided tour for high dimensional visualisation.

The rest of this paper is structured as follows. @sec-background
introduces the background of the PP guided tour and reviews existing methods in the literature. @sec-theory describes the details of JSO
and introduces metrics that measure different properties of a PPI.
@sec-vis visualises the behaviour of the JSO in terms of describes a simulation study on the performance of the JSO using well-known data sets and index functions.
@sec-app displays two sets of simulation experiments comparing the JSO with the search-better optimiser and studies the impact of different PPI properties on the optimisation performance. @sec-conclusion summarises the work and provides suggestions for future directions.

@@

The speed of optimizing an index function can be measured empirically by recording the duration of the optimisation.
Alternatively, one can also gauge the speed in terms of computational complexity of evaluating the index function (for instance, in big O notation, with respect to
the sample size) of computing the index function.

### 2024-06-20 optim.tex
intro

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{algorithm}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Multivariate Analysis}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num-names}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Wai Yin Leung},
  pdfkeywords={projection pursuit, Jellyfish Search Optimiser
(JSO), Optimisation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    projection pursuit \sep Jellyfish Search Optimiser (JSO) \sep 
    Optimisation
\end{keyword}
\end{frontmatter}

@@

\section{Introduction {[}Nicolas and
Jessica{]}}\label{introduction-nicolas-and-jessica}

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely \emph{projection
pursuit index (PPI)}, (\citet{hall1989polynomial},
\citet{cook1993projection}, \citet{lee2010projection},
\citet{Loperfido2018}, \citet{Loperfido2020}). Given high-dimensional
data \(X \in \mathcal{R}^{n }\times p\) and PPI \(f(\cdot)\), PP finds
the orthonormal projection matrix \(A \in \mathcal{R}^{p \times d}\) by
solving the following optimisation problem:

\[
\underset{A}{\max} f(XA) \quad \text{subject to} \quad A'A = I
\]

@@

The artificial jellyfish search (JS) algorithm \citep{chou_novel_2021}
is a swarm-based metaheuristic optimisation algorithm inspired by the
search behaviour of jellyfish in the ocean. It is one of the latest
swarm intelligence algorithms \citep{rajwar_exhaustive_2023}, which was
shown to have stronger search ability and faster convergence with fewer
parameters compared to classic optimisation methods
\citep{chou_novel_2021}-\citep{chou_recent_2022}.

The primary goal of this study is to investigate the performance of JSO
in the context of the PP guided tour. A series of simulation experiments
using well-recognised data sets and PPIs are conducted to yield insights
regarding the behaviour of JSO under different settings and the
sensitivity of hyper-parameter in the optimisation. Second, to observe
the performance of JSO with different types of PPI, a collection of
metrics is introduced to capture specific properties of the index
including squintability and smoothness as introduced in
\citet{laa_using_2020}. To the best of the author's knowledge, this is
the first attempt to quantitatively measure squintability and
smoothness. Finally, the relationship between the JSO performance,
hyper-parameters tuning and various properties of PPI is analysed to
provide helpful guidance for practitioners that are using the guided
tour for high dimensional visualisation.

The rest of this paper is structured as follows.
Section~\ref{sec-background} introduces the PP guided tour.
Section~\ref{sec-theory} describes the details of JSO and introduces
metrics that measure different characteristics of a PPI.
\textbf{?@sec-simulation} describes a simulation study on the
performance of the JSO using well-known data sets and index functions.
Section~\ref{sec-conclusion} summarises the work and provides
suggestions for future directions.

\section{Projection pursuit, index functions and optimisation {[}Di and
Sherry{]}}\label{sec-background}

@@

\begin{tcolorbox}[enhanced jigsaw, colback=white, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toptitle=1mm, opacityback=0, toprule=.15mm, leftrule=.75mm, coltitle=black, bottomtitle=1mm, arc=.35mm, rightrule=.15mm, colbacktitle=quarto-callout-note-color!10!white, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, opacitybacktitle=0.6, left=2mm, breakable]

@@

\subsection{Smoothness}\label{sec-smoothness}

@@

\subsection{Squintability}\label{sec-squintability}

@@

\subsection{Speed}\label{speed}

The speed of optimizing an index function can be calculated/measured
using the computational complexity (in big O notation, with respect to
the sample size) of computing the index function.

\section{Visualisation of jellyfish optimiser}\label{sec-vis}

@@

\centering{

\includegraphics{optim_files/figure-pdf/fig-projection-1.pdf}

@@

\end{figure}%

@@

\centering{

\includegraphics{optim_files/figure-pdf/fig-index-value-proj-dist-1.pdf}

@@

\end{figure}%

\section{Application {[}Di and Sherry{]}}\label{sec-app}

@@

\subsection{Performance of the JS optimiser in pipe-finding
problems}\label{sec-app-1}

@@

\centering{

\includegraphics{optim_files/figure-pdf/fig-proj-1.pdf}

@@

\centering{

\includegraphics{optim_files/figure-pdf/fig-proportion-1.pdf}

@@

\subsection{Factors affecting the JS optimiser success rate:
optimisation properties and jellyfish hyper-parameters}\label{sec-app-2}

@@

\centering{

\includegraphics{optim_files/figure-pdf/fig-idx-proj-dist-1.pdf}

@@

\begin{table}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (\(\theta_1\) to \(\theta_4\)) for the
pipe-finding and sine-wave finding problems. The column ``smooth'' and
``squint'' represent the smoothness and squintability measures.}

\centering{

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrr>{}r|>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 0.763\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 0.740\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 0.737\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 0.779\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 0.314\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.193\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 0.330\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 0.342\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.737\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 0.689\\
\addlinespace
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 0.780\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 0.739\\
\bottomrule
\end{tabular}
\endgroup{}

\end{table}%

@@

\begin{longtable}[t]{lrrrrrrl}

\caption{\label{tbl-mod-data}The first 7 rows of the datasets processed
for modelling.}

\tabularnewline

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 0.314 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 0.314 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 0.314 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 0.314 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 0.314 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.193 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.193 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

\end{longtable}

\endgroup{}

@@

\begin{longtable}[t]{|>{}lrrr>{}r|}

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success on optimisation properties and jellyfish
hyper-parameters from the generalised linear model with a binomial
family and a logit link function. The variable smoothness,
squintability, number of jellyfish and maximum number of tries are
positively associated with the JS optimiser success rate while data
dimension and being flagged as long runtime are negatively associated
with the success rate. The variable squintability and dimension are
significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success.}

\tabularnewline

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 8.241 & 2.737 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule

\end{longtable}

\endgroup{}

\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{algorithm}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal of Multivariate Analysis}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num-names}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit},
  pdfauthor={H. Sherry Zhang; Dianne Cook; Nicolas Langrené; Jessica Wai Yin Leung},
  pdfkeywords={projection pursuit, Jellyfish Search Optimiser
(JSO), Optimisation},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{keyword}
    projection pursuit \sep Jellyfish Search Optimiser (JSO) \sep 
    Optimisation
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, frame hidden, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{introduction-nicolas-and-jessica}{%
\section{Introduction {[}Nicolas and
Jessica{]}}\label{introduction-nicolas-and-jessica}}

Projection Pursuit (PP) is a high-dimensional data visualisation
approach that involves computing informative linear projections of data
by optimising a variety of objective functions, namely \emph{projection
pursuit index (PPI)} (\citet{hall1989polynomial},
\citet{cook1993projection}, \citet{lee2010projection},
\citet{Loperfido2018}, \citet{Loperfido2020}). Given high-dimensional
data \(X \in \mathbf{R}^{n }\times p\) and PPI \(f(\cdot)\), PP finds
the orthonormal projection matrix \(A \in \mathbf{R}^{p \times d}\) by
solving the following optimisation problem:

\[
\underset{A}{\max } \quad f(XA) \quad \text{subject to} \quad A'A = I
\]

@@

The artificial Jellyfish Search Optimiser (JSO) \citep{chou_novel_2021}
is a swarm-based metaheuristic optimisation algorithm inspired by the
search behaviour of jellyfish in the ocean. It is one of the latest
swarm intelligence algorithms \citep{rajwar_exhaustive_2023}, which was
shown to have stronger search ability and faster convergence with fewer
parameters compared to classic optimisation methods
\citep{chou_novel_2021}-\citep{chou_recent_2022}.

The primary goal of this study is to investigate the performance of JSO
in the context of the PP guided tour. A series of simulation experiments
using well-recognised data sets and PPIs are conducted to yield insights
regarding the behaviour of JSO under different settings and the
sensitivity of hyper-parameter in the optimisation. Second, to observe
the performance of JSO with different types of PPI, a collection of
metrics is introduced to capture specific properties of the index
including squintability and smoothness as introduced in
\citet{laa_using_2020}. To the best of the authors' knowledge, this is
the first attempt to quantitatively measure squintability and
smoothness. Finally, the relationship between the JSO performance,
hyper-parameters tuning and various properties of PPI is analysed to
provide helpful guidance for practitioners that are using the guided
tour for high dimensional visualisation.

The rest of this paper is structured as follows.
Section~\ref{sec-background} introduces the background of the PP guided
tour and reviews existing methods in the literature.
Section~\ref{sec-theory} describes the details of JSO and introduces
metrics that measure different properties of a PPI.
Section~\ref{sec-vis} visualises the behaviour of the JSO in terms of
describes a simulation study on the performance of the JSO using
well-known data sets and index functions. Section~\ref{sec-app} displays
two sets of simulation experiments comparing the JSO with the
search-better optimiser and studies the impact of different PPI
properties on the optimisation performance. Section~\ref{sec-conclusion}
summarises the work and provides suggestions for future directions.

\hypertarget{sec-background}{%
\section{Projection pursuit, index functions and optimisation {[}Di and
Sherry{]}}\label{sec-background}}

@@

\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, leftrule=.75mm, left=2mm, colbacktitle=quarto-callout-note-color!10!white, colback=white, arc=.35mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toptitle=1mm, opacityback=0, bottomtitle=1mm, breakable, rightrule=.15mm, bottomrule=.15mm, titlerule=0mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, coltitle=black]

@@

\hypertarget{sec-smoothness}{%
\subsection{Smoothness}\label{sec-smoothness}}

@@

\hypertarget{sec-squintability}{%
\subsection{Squintability}\label{sec-squintability}}

@@

\hypertarget{speed}{%
\subsection{Speed}\label{speed}}

The speed of optimizing an index function can be measured empirically by
recording the duration of the optimisation. Alternatively, one can also
gauge the speed in terms of computational complexity of evaluating the
index function (for instance, in big O notation, with respect to the
sample size) of computing the index function.

\hypertarget{sec-vis}{%
\section{Visualisation of jellyfish optimiser}\label{sec-vis}}

@@

{\centering \includegraphics{optim_files/figure-pdf/fig-projection-1.pdf}

@@

\end{figure}

@@

{\centering \includegraphics{optim_files/figure-pdf/fig-index-value-proj-dist-1.pdf}

@@

\end{figure}

\hypertarget{sec-app}{%
\section{Application {[}Di and Sherry{]}}\label{sec-app}}

@@

\hypertarget{sec-app-1}{%
\subsection{Performance of the JS optimiser in pipe-finding
problems}\label{sec-app-1}}

@@

{\centering \includegraphics{optim_files/figure-pdf/fig-proj-1.pdf}

@@

{\centering \includegraphics{optim_files/figure-pdf/fig-proportion-1.pdf}

@@

\hypertarget{sec-app-2}{%
\subsection{Factors affecting the JS optimiser success rate:
optimisation properties and jellyfish
hyper-parameters}\label{sec-app-2}}

@@

{\centering \includegraphics{optim_files/figure-pdf/fig-idx-proj-dist-1.pdf}

@@

\hypertarget{tbl-smoothness-squintability}{}
\begin{table}
\caption{\label{tbl-smoothness-squintability}Parameters estimated from the Gaussian process (including variance,
range, smoothness, and nugget) and scaled logistic function
(\(\theta_1\) to \(\theta_4\)) for the pipe-finding and sine-wave
finding problems. The column ``smooth'' and ``squint'' represent the
smoothness and squintability measures. }\tabularnewline

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrr>{}r|>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 0.763\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 0.740\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 0.737\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 0.779\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 0.314\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.193\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 0.330\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 0.342\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.737\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 0.689\\
\addlinespace
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 0.780\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 0.739\\
\bottomrule
\end{tabular}
\endgroup{}
\end{table}

@@

\hypertarget{tbl-mod-data}{}
\begin{longtable}[t]{lrrrrrrl}
\caption{\label{tbl-mod-data}The first 7 rows of the datasets processed for modelling. }\tabularnewline

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 0.314 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 0.314 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 0.314 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 0.314 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 0.314 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.193 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.193 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule
\end{longtable}
\endgroup{}

@@

\hypertarget{tbl-mod-output}{}
\begin{longtable}[t]{|>{}lrrr>{}r|}
\caption{\label{tbl-mod-output}Model estimates of proportion of jellyfish success on optimisation
properties and jellyfish hyper-parameters from the generalised linear
model with a binomial family and a logit link function. The variable
smoothness, squintability, number of jellyfish and maximum number of
tries are positively associated with the JS optimiser success rate while
data dimension and being flagged as long runtime are negatively
associated with the success rate. The variable squintability and
dimension are significant, suggesting their importance relative to
jellyfish hyper-parameters in the optimisation success. }\tabularnewline

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 8.241 & 2.737 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule
\end{longtable}
\endgroup{}

\hypertarget{sec-conclusion}{%
\section{Conclusion {[}Di and Sherry{]}}\label{sec-conclusion}}
